\chapter{Conclusion}\label{ch:conclusion}

Evaluating the readability and ``human-likeness'' of decompiled code has long been a subjective and labor-intensive challenge in reverse engineering. This master's thesis explored the viability of using \acp{LLM} to automate this evaluation, investigating both statistical metrics (Perplexity) and qualitative prompting strategies (LLM-as-a-Judge).

Our quantitative analysis definitively demonstrated that average perplexity is not a reliable proxy for human-likeness in decompiled code. Contrary to initial assumptions, human-written source code consistently exhibits higher perplexity than decompiled output. This is because human code contains domain-specific idioms, creative structural choices, and stylistic variance (high entropy), whereas decompilers generate rigid, repetitive, and verbose boilerplate that artificially drives perplexity down. Thus, the correlation between $\Delta$perplexity and human preference was weak and often inversely related, confirming that perplexity alone cannot capture the qualities that make code more readable or natural to human developers.

To address the shortcomings of average perplexity, we investigated the use of Total Loss as an alternative quantitative metric. By accumulating prediction errors across all tokens, Total Loss inherently penalizes structural bloat and verbosity. Our preliminary findings suggest that Total Loss, particularly when computed on the anonymized \ac{AST} representation to remove decompiler-specific lexical noise, correlates much more strongly with both \ac{LLM} preferences and human judgments. While these initial results are promising and indicate that \ac{AST} Total Loss could serve as a better proxy for structural code quality, using less resources and time respect the LLM-as-a-Judge approach, it should not be viewed as a definitive solution yet; further in-depth studies are required to fully validate its robustness and generalizability.

In exploring the LLM-as-a-Judge paradigm, our results highlighted distinct capabilities and vulnerabilities between the evaluated models. \emph{Deepseek-r1} emerged as a highly capable evaluator, achieving an alignment of approximately 74\% with human developers. We demonstrated that abstracting the decompiled code into an \ac{AST} successfully mitigated some lexical distractions, forcing the models to evaluate purely structural logic and improving their alignment with human reasoning.

Despite these successes, we identified significant limitations. \acp{LLM} remain susceptible to analytical hallucinations, occasionally inventing non-existent technical justifications when evaluating logically equivalent snippets or minor stylistic permutations (such as reordering \texttt{switch} cases). Furthermore, the computational cost of processing long contexts for perplexity and loss calculations poses a scalability bottleneck for current hardware.

\section{Limitations}

This study has several limitations that should be acknowledged:
\begin{itemize}
    \item \textbf{Model size and architecture:} The \acp{LLM} evaluated in this study are relatively small compared to state-of-the-art models. Larger models with more parameters may have different capabilities and biases, which could affect the generalizability of our findings.
    \item \textbf{Dataset size and diversity:} The dataset of decompiled code snippets used for evaluation is limited in size and may not fully represent the diversity of real-world decompilation scenarios. A larger and more varied dataset could provide a more comprehensive assessment of \ac{LLM} performance.
    \item \textbf{Human validation sample size:} The human survey conducted to establish a ground truth for readability preferences involved a small number of participants evaluating a limited subset of 21 questions. Consequently, while the survey provides valuable preliminary insights, particularly regarding the correlation with \ac{AST} Total Loss, it lacks the statistical power required for definitive conclusions.
    \item \textbf{Prompt engineering:} The design of prompts can significantly influence \ac{LLM} responses. While we attempted to create balanced and informative prompts, different prompt formulations could yield different results, introducing variability in the evaluation.
    \item \textbf{Human evaluation subjectivity:} The human judgments used as a benchmark for evaluating \ac{LLM} performance are inherently subjective. Different evaluators may have varying opinions on what constitutes more readable or human-like code, which can affect the consistency of the evaluation.
    \item \textbf{Fixed base:} In our evaluation, we compared PRs against a fixed base version. This approach does not isolate the specific changes made in some older \acp{PR} but rather evaluates different versions of the same codebase. Future work could focus on more controlled experiments that isolate specific transformations to better understand their impact on readability.
\end{itemize}

In conclusion, while \acp{LLM} are not yet flawless, objective arbiters of code quality, they represent a powerful and promising tool for the reverse engineering community. 
The findings of this study establish a solid baseline for automated decompiler evaluation. 
Future work should focus on utilizing larger foundation models, using much bigger datasets, and engaging a significantly larger pool of developer judges to rigorously validate the Total Loss metric and refine the LLM-as-a-Judge approach, while also exploring techniques to mitigate hallucinations and reduce computational overhead.

\chapter*{Acknowledgements}

I would like to express my heartfelt thanks to all those who have supported and encouraged me throughout my academic career. First and foremost, I would like to thank my parents and my family for their love, patience, and unconditional support.
I would also like to thank my groups of friends,
the \emph{ThaSuS} group, the \emph{Office} group, the University Group of bachelor's and master's degree students, and all my other friends scattered here and there whom I would like to see more often, for the laughter, the discussions, the good times, and the mutual support we have given each other.
Finally, I would like to thank my advisors and all the professors who have guided me throughout my university career
for their expertise, availability, and valuable advice. Without the support of all these
people, I would not have made it this far. Thank you all from the bottom of my heart!
